### 2019-11-15 s1_text/s1_text.tex
MJM additions to S1 supplement on memb transporter details

\usepackage[obeyFinal]{todonotes} % todo notes and missing figures

@@

\begin{flushleft}
	{\Large
		\textbf\newline{Supporting information for ``theoretical investigation of a genetic switch for metabolic adaptation''}
	}
	\newline
	% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
	\\
	Kathrin Laxhauber\textsuperscript{1},
	Muir J. Morrison\textsuperscript{2},
	Rob Phillips\textsuperscript{2,3$\ast$}
	\\
	\bigskip
	\textbf{1} Dept???, ETHZ, Zurich, State???, Switzerland\todo{MJM: Add dept \& complete affiliation}
	\\
	\textbf{2} Department of Physics, California Institute of Technology, Pasadena, CA, USA
	\\
	\textbf{3} Department of Biology and Biological Engineering, California Institute of Technology, Pasadena, CA, USA
	\\
	\bigskip

	% Insert additional author notes using the symbols described below. Insert symbol callouts after author names as necessary.
	% 
	% Remove or comment out the author notes below if they aren't used.
	%
	% Primary Equal Contribution Note
	% Yinyang These authors contributed equally to this work.

	% Additional Equal Contribution Note
	% Also use this double-dagger symbol for special authorship notes, such as senior authorship.
	% \ddag These authors also contributed equally to this work.

	% Current address notes
	% \textcurrency Current Address: Dept/Program/Center, Institution Name, City, State, Country % change symbol to "\textcurrency a" if more than one current address note
	% \textcurrency b Insert second current address 
	% \textcurrency c Insert third current address

@@

\paragraph*{Kinetics of XapB, NupC, and NupG.}
All transporters in our system are driven by the proton
gradient~\cite{Norholm2001}, and we assume their kinetic scheme to be like
that of \emph{lac} permease. This scheme is shown schematically in
Fig~\ref{figS2:permease}. Even though it includes several steps, the
essential part is substrate binding, conformational change of the
transporter, and substrate release on the other side of the membrane. We
assume the proton gradient across the membrane to be constant and proton
binding and unbinding to the transporter to be fast. Hence, the latter does
not need to be included explicitly in the kinetic model (it is implicitly
part of the turnover rate).

@@

The NupC and NupG reactions are approximated as linearly dependent on the
substrate. This should be appropriate, since their affinity for xanthosine
is so low that it cannot even be measured properly~\cite{Norholm2001}. This
makes $K_{\n{M}}$ much larger than $c$ or $\n{[x]}$ will ever be. For XapB,
this is not the case, because it has much higher specificity for xanthosine
and $c$ and $\n{[x]}$ take values around the $K_{\n{M}}$ values of XapB.

@@

To find a lower bound, we estimate the diffusion rate. As described in
Eq~\ref{eq:diff} in the previous section, the diffusion rate can be written
as $10^{-2} \cdot p \cdot A$, where $p$ is the permeability coefficient of
the membrane for xanthosine, $A$ is the total cell surface area, and
$10^{-2}$ is the fraction of uncharged xanthosine at physiological pH. From
the permeabilities of various substances that are collected
in~\cite{Milo2016} and from the measured permeability values for purine
nucleosides in~\cite{Wei2011}, we estimate a permeability coefficient of
$\tilde{p}=10^{-8 \pm 3}~\n{m\ s^{{-1}}}$. We write the tilde because this
value is given in a convention where $\tilde{p} \cdot A = \dd{x}{t}$ with
$x$ being the number of xanthosine molecules and not the concentration.
Since we work in concentrations, we divide by an estimated cell volume of
$10^{-18}~\n{m^3}$ and, with a cell surface area of $10^{-11}~\n{m^2}$, find
a diffusion rate of roughly $10^{-3 \pm 3}~\n{s^{-1}}$. This sets the lower
bound of $k_{\n{nup}}$.
%\com{Check if Overton's rule gives the samediffusion coefficient}

@@

\paragraph*{Hysteresis effects.}
In Fig~\ref{figS8:hysteresis}, the results mentioned in the hysteresis part
of the main text are shown. As mentioned there,  the simulation was started
with initially fully induced cells. At concentrations where no cells
switched from the low to the high expression state before, cells here do not
switch back to low expression.

>>>>

@@

\begin{flushleft}
	{\Large
		\textbf\newline{Supporting information for ``theoretical investigation of a genetic switch for metabolic adaptation''}
	}
	\newline
	% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
		\\
	Kathrin S. Laxhuber\textsuperscript{1},
	Muir J. Morrison\textsuperscript{2},
	Rob Phillips\textsuperscript{2,3$\ast$}
	\\
	\bigskip
	\textbf{1} Department of Chemistry and Applied Biosciences, ETH Zurich, 8093 Zurich, Switzerland
	\\
	\textbf{2} Department of Physics, California Institute of Technology, Pasadena, CA, USA
	\\
	\textbf{3} Department of Biology and Biological Engineering, California Institute of Technology, Pasadena, CA, USA
	\\
	\bigskip

	% Insert additional author notes using the symbols described below. Insert symbol callouts after author names as necessary.
	% 
	% Remove or comment out the author notes below if they aren't used.
	%
	% Primary Equal Contribution Note
	%\Yinyang These authors contributed equally to this work.

	% Additional Equal Contribution Note
	% Also use this double-dagger symbol for special authorship notes, such as senior authorship.
	%\ddag These authors also contributed equally to this work.

	% Current address notes
	%\textcurrency Current Address: Dept/Program/Center, Institution Name, City, State, Country % change symbol to "\textcurrency a" if more than one current address note
	% \textcurrency b Insert second current address 
	% \textcurrency c Insert third current address

	% Deceased author note
	%\dag Deceased

	% Group/Consortium Author Note
	%\textpilcrow Membership list can be found in the Acknowledgments section.

@@

\paragraph*{Kinetics of XapB, NupC, and NupG.}
Entering the cell consists of two steps. In a first transport step, xanthosine passes the
outer membrane through porins like Tsx, OmpF and OmpC. This does not seem to
be a rate-limiting step. Then, XapB actively transports it across the inner
membrane, powered by the proton motive force. There are two other active
nucleoside transporters that seem to be transporting xanthosine with a very
low affinity: NupC and NupG. It was found that \textDelta\emph{nupC}
\textDelta\emph{nupG} strains cannot grow on xanthosine. Hence, these seem
to be necessary to ``start'' the system by importing small amounts of
xanthosine that then activate \emph{xapB}, and there appears to be no other
significant way in which xanthosine can enter the cell. All statements in this paragraph are based on data
from~\cite{Norholm2001}.

As mentioned in the main text, we assume the kinetic scheme of all three transporters to be similar to that
of \emph{lac} permease (as it is described in~\cite{Kaback2015}), which is discussed in more detail in the next subsection. One transporter of this type can do two things to change the intracellular xanthosine
concentration $\n{[x]}$: it can actively transport one xanthosine into the
cell, powered by the proton gradient, or it can transport one out of the
cell, against the proton gradient. For net transport, a proton and a
substrate need to bind on one side of the membrane and detach from the
transporter on the other side of the membrane before the transporter goes
back to the first side again (see next subsection). Because of the
proton gradient, influx is overall thermodynamically more likely than export
at low intracellular xanthosine concentrations, which leads to a net influx.
For much higher intra- than extracellular xanthosine concentrations, the
difference in the chemical potential of xanthosine across the membrane can
dominate that of the protons and there is net efflux.

We model influx and efflux separately and use Michaelis-Menten kinetics for
XapB (this is discussed in more detail in the next subsection). For the turnover rate and Michaelis constant for influx we write
$k_{\n{b, i}}$ and $K_{\n{b, i}}$, respectively, and $k_{\n{b, e}}, K_{\n{b,
		e}}$ for efflux. These parameters implicitly include the proton gradient.

For NupC and NupG, we approximate the transport reactions as linearly dependent on the substrate. This should be appropriate, since their affinity for xanthosine is so low that it cannot even be measured properly~\cite{Norholm2001}, which makes $K_{\n{M}}$ much larger than $c$ or $\n{[x]}$ will ever be. Thus, the general $k_{\n{cat}} \n{[E]_0}
\frac{\n{[S]}}{K_{\n{M}} + \n{[S]}}$ can be approximated as
$\frac{k_{\n{cat}}}{K_{\n{M}}} \n{[E]_0} \n{[S]}$. In our equations,
$\frac{k_{\n{cat}}}{K_{\n{M}}} \n{[E]_0}$ is denoted by $k_{\n{nup}}$ for
influx, and the efflux rate is written as $\xi k_{\n{nup}}$. For the XapB
kinetics, no such approximation can be made, because the xanthosine
concentration in the dynamic system can range from far below the respective
$K_{\n{M}}$ value to far above.

\paragraph*{The kinetics of \emph{lac} permease.}
All transporters in our system are driven by the proton gradient~\cite{Norholm2001}, and we assume their kinetic scheme to be like that of \emph{lac} permease. This scheme is shown schematically in Fig~\ref{figS2:permease}.
Even though it includes several steps, the essential part is substrate binding, conformational change of the transporter, and substrate release on the other side of the membrane. We assume the proton gradient across the membrane to be constant and proton binding and unbinding to the transporter to be fast. Hence, the latter does not need to be included explicitly in the kinetic model (it is implicitly part of the turnover rate).

@@

@@

To find a lower bound, we estimate the diffusion rate. As described in Eq~\ref{eq:diff} in the previous section, the diffusion rate can be written as $10^{-2} \cdot p \cdot A$, where $p$ is the permeability coefficient of the membrane for xanthosine, $A$ is the total cell surface area, and $10^{-2}$ is the fraction of uncharged xanthosine at physiological pH. From the permeabilities of various substances that are collected in~\cite{Milo2016} and from the measured permeability values for purine nucleosides in~\cite{Wei2011}, we estimate a permeability coefficient of $\tilde{p}=10^{-8 \pm 3}~\n{m\ s^{{-1}}}$. We write the tilde because this value is given in a convention where $\tilde{p} \cdot A = \dd{x}{t}$ with $x$ being the number of xanthosine molecules and not the concentration. Since we work in concentrations, we divide by an estimated cell volume of $10^{-18}~\n{m^3}$ and, with a cell surface area of $10^{-11}~\n{m^2}$, find a diffusion rate of roughly $10^{-3 \pm 3}~\n{s^{-1}}$. This sets the lower bound of $k_{\n{nup}}$. %\com{Check if Overton's rule gives the same diffusion coefficient}

@@

\paragraph*{Hysteresis effects.}
In Fig~\ref{figS8:hysteresis}, the results mentioned in the hysteresis part of the main text are shown. As mentioned there,  the simulation was started with initially fully induced cells. At concentrations where no cells switched from the low to the high expression state before, cells here do not switch back to low expression. The extracellular xanthosine concentration (for the other parameters being as in the Table in the main text) needed to be as low as $\n{[c]_a} = 10$ to observe bimodality and $\n{[c]_a}=9$ to have all cells in the low expression state (recall $\n{[c]_{a}} \defeq \frac{c}{K_{\n{a}}}$, $K_{\n{a}} = 5 \cdot
10^{-5} \unit{M}$).
